---
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment of Course5week2

## Loading and Preprocessing the Data
First the data was downloaded, unzipped and then read. After that the date variable was converted to date class.

```{r loading code,cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "./course5week2.zip")
data<-unzip("course5week2.zip")
data <- read.csv(data)
data$date<-as.Date(as.character(data$date),"%Y-%m-%d")
```

## Mean Total Number of Steps Taken Per Day With Missing Values Ignored

Getting data set with no NA values and then summing the steps per days

```{r number of steps taken per day, cache=TRUE}
dataNoNa<-data[!(is.na(data$steps)),] #getting dataSet with no Na values
totalStepsPerDay<-aggregate(steps~date,FUN=sum,data = dataNoNa) #summing data
```
Plotting the total steps per each day

```{r,cache=TRUE}
library(ggplot2)
p<-ggplot(data=totalStepsPerDay,aes(x=date,y=steps))
p+geom_col() #plotting 
```
Calculating Mean and Median of the total number of steps taken per day

```{r,cache=TRUE}
totalMeanStepsPerDay<-aggregate(steps~date,FUN=mean,data = dataNoNa) #getting mean of steps per days
totalMedianStepsPerDay<-aggregate(steps~date,FUN=median,data = dataNoNa) #getting median of steps per days
meanAndMedian<-merge(totalMeanStepsPerDay,totalMedianStepsPerDay,by="date") #merging data by means of date
names(meanAndMedian)<-c("Date","mean of Steps","median of steps") # changing name
meanAndMedian
```

##  The Average Daily Activity Pattern

Now we will average all steps per each 5 mint interval

```{r,cache=TRUE}
totalStepsPertime<-aggregate(steps~interval,FUN=mean,data = dataNoNa) #For Making a time series plotof the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
plot(totalStepsPertime$interval,totalStepsPertime$steps,type = "l",xlab = "Interval changing by 5 mint",ylab = "Steps")
totalStepsPertime[totalStepsPertime$steps==max(totalStepsPertime$steps),] #showing at which interval maximum steps occur

```

## Imputing Missing Values
 First shwoing the no of rows having missing values
```{r missing rows,cache=TRUE}
dataWithNa<-data[is.na(data$steps),]
missingValues<-nrow(dataWithNa) #showing missing Values
missingValues
```

To compute the missing values, average of all days was taken on each 5 mint interval from the data having no missng values. Then this average steps were included in the Na values of the missing data with the correct time interval

```{r imputing values,cache=TRUE,results="hide"}
totalStepsPertime<-aggregate(steps~interval,FUN=mean,data = dataNoNa)
includingDataForNa<-merge(totalStepsPertime,dataWithNa,by="interval")
includingDataForNa$steps.y<-NULL # deleting the steps.y variable.
colnames(includingDataForNa)[2]<-"steps"  #changing name of the variable
completeData<-rbind(includingDataForNa,dataNoNa)  #combining data of cpmute values and predata values
library(dplyr)
completeData<-arrange(completeData,date) # complete data arranged by date
```
A histogram of the total number of steps taken each day and Calculating and reporting the mean and median total number of steps taken per day

```{r histogram,cache=TRUE}
totalStepsPerDayx<-aggregate(steps~date,FUN=sum,data = completeData) #computing total steps per day for the new data
q<-ggplot(data=totalStepsPerDayx,aes(x=date,y=steps))
q+geom_col() #plotting histogram
totalMeanStepsPerDayx<-aggregate(steps~date,FUN=mean,data = completeData) #getting mean of steps per days
totalMedianStepsPerDayx<-aggregate(steps~date,FUN=median,data = completeData) #getting median of steps per days
meanAndMedianx<-merge(totalMeanStepsPerDayx,totalMedianStepsPerDay,by="date") #merging data by means of date
names(meanAndMedianx)<-c("Date","mean of Steps","median of steps") # changing name
meanAndMedianx
```
By following the mean method for imputing values there is no change in the mean and median when compare to the mean and median calculated from Na values removed data.

```{r change in calculation,cache=TRUE}
changeInmean<-meanAndMedian$`mean of Steps`- meanAndMedianx$`mean of Steps`
table(changeInmean)
changeInMedian <-meanAndMedian$`median of steps` - meanAndMedianx$`median of steps`
table(changeInMedian)
```
## Seeing Differences In Activity Patterns Between Weekdays and Weekends
Creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r for new factor variable,cache=TRUE}
weekdaysOfMonth <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday') #creating a factor variable of weekday and weekend
completeData$whichDay <- factor((weekdays(completeData$date) %in% weekdaysOfMonth), 
                   levels=c(FALSE, TRUE), labels=c('weekend', 'weekday') )
```

Making a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r code for panel plot for weekdays and weekend, cache=TRUE}
meanWrtIntervalAndDays<-aggregate(steps~interval+whichDay,FUN=mean,data = completeData) #arranging data with respect to days and interval
qplot(interval,steps,data = meanWrtIntervalAndDays,geom = c("line"),facets = .~whichDay)
```

